---
title: "Analyzing single-cell RNA-seq data with Bioconductor (UMI counts)"
author: 
- name: Aaron T. L. Lun
  affiliation: &CRUK Cancer Research UK Cambridge Institute, Li Ka Shing Centre, Robinson Way, Cambridge CB2 0RE, United Kingdom
- name: Davis J. McCarthy
  affiliation: 
  - &EMBL EMBL European Bioinformatics Institute, Wellcome Genome Campus, Hinxton, Cambridge CB10 1SD, United Kingdom
  - St Vincent's Institute of Medical Research, 41 Victoria Parade, Fitzroy, Victoria 3065, Australia
- name: John C. Marioni
  affiliation: 
  - *CRUK
  - *EMBL
  - Wellcome Trust Sanger Institute, Wellcome Genome Campus, Hinxton, Cambridge CB10 1SA, United Kingdom
date: 27 October 2017
vignette: >
  %\VignetteIndexEntry{A worfklow for low-level analyses of single-cell RNA-seq data
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}    
output: 
    BiocStyle::html_document
bibliography: ref.bib
---

```{r style, echo=FALSE, results='hide', message=FALSE}
library(BiocStyle)
library(knitr)
opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)

# Setting single-core unless explicitly specified otherwise.
library(BiocParallel)
register(SerialParam())

# Deciding whether we want to re-download everything or not.
on.bioc <- FALSE

# Further arguments for local execution.
opts_chunk$set(fig.asp=1)
if (!on.bioc) {
    opts_chunk$set(dpi=300, dev="png", dev.args=list(pointsize=15))
    options(bitmapType="cairo", width=100)
}

# Additional required libraries.
library(Rtsne)
```

```{r, eval=on.bioc, echo=FALSE, results='hide'}
all.urls <- c("https://storage.googleapis.com/linnarsson-lab-www-blobs/blobs/cortex/expression_mRNA_17-Aug-2014.txt",
"https://storage.googleapis.com/linnarsson-lab-www-blobs/blobs/cortex/expression_mito_17-Aug-2014.txt",
"https://storage.googleapis.com/linnarsson-lab-www-blobs/blobs/cortex/expression_spikes_17-Aug-2014.txt")
for (x in seq_along(all.urls)) { 
    download.file(all.urls[x], all.basenames[x], mode=all.modes[x])
}
```

# Overview

We examine a heterogeneous dataset from a study of cell types in the mouse brain [@zeisel2015brain].
This contains approximately 3000 cells of varying types such as oligodendrocytes, microglia and neurons.
Individual cells were isolated using the Fluidigm C1 microfluidics system and library preparation was performed on each cell using a UMI-based protocol.
After sequencing, expression was quantified by counting the number of UMIs mapped to each gene.
Count data for all endogenous genes, mitochondrial genes and spike-in transcripts were obtained from http://linnarssonlab.org/cortex.

# Count loading 

The count data are distributed across several files, so some work is necessary to consolidate them into a single matrix.
We define a simple utility function for loading data in from each file. 
(We stress that this function is only relevant to the current dataset, and should not be used for other datasets.
This kind of effort is generally not required if all of the counts are in a single file and separated from the metadata.)

```{r}
readFormat <- function(infile) { 
    # First column is empty.
    metadata <- read.delim(infile, stringsAsFactors=FALSE, header=FALSE, nrow=10)[,-1] 
    rownames(metadata) <- metadata[,1]
    metadata <- metadata[,-1]
    metadata <- as.data.frame(t(metadata))

    # First column after row names is some useless filler.
    counts <- read.delim(infile, stringsAsFactors=FALSE, 
        header=FALSE, row.names=1, skip=11)[,-1] 
    counts <- as.matrix(counts)
    return(list(metadata=metadata, counts=counts))
}
```

Using this function, we read in the counts for the endogenous genes, ERCC spike-ins and mitochondrial genes.

```{r}
endo.data <- readFormat("expression_mRNA_17-Aug-2014.txt")
spike.data <- readFormat("expression_spikes_17-Aug-2014.txt")
mito.data <- readFormat("expression_mito_17-Aug-2014.txt")
```

We also need to rearrange the columns for the mitochondrial data, as the order is not consistent with the other files.

```{r}
m <- match(endo.data$metadata$cell_id, mito.data$metadata$cell_id)
mito.data$metadata <- mito.data$metadata[m,]
mito.data$counts <- mito.data$counts[,m]
```

```{r, echo=FALSE}
stopifnot(identical(endo.data$metadata$cell_id, spike.data$metadata$cell_id)) # should be the same.
stopifnot(all(endo.data$metadata$cell_id==mito.data$metadata$cell_id)) # should now be the same.
```

In this particular data set, some genes are represented by multiple rows corresponding to alternative genomic locations.
We sum the counts for all rows corresponding to a single gene for ease of interpretation.

```{r}
raw.names <- sub("_loc[0-9]+$", "", rownames(endo.data$counts))
new.counts <- rowsum(endo.data$counts, group=raw.names, reorder=FALSE)
endo.data$counts <- new.counts
```

The counts are then combined into a single matrix for constructing a `SingleCellExperiment` object.
For convenience, metadata for all cells are stored in the same object for later access.

```{r}
library(SingleCellExperiment)
all.counts <- rbind(endo.data$counts, mito.data$counts, spike.data$counts)
sce <- SingleCellExperiment(list(counts=all.counts), colData=endo.data$metadata)
dim(sce)
```

We also add annotation identifying rows that correspond to each class of features.

```{r}
nrows <- c(nrow(endo.data$counts), nrow(mito.data$counts), nrow(spike.data$counts))
is.spike <- rep(c(FALSE, FALSE, TRUE), nrows)
is.mito <- rep(c(FALSE, TRUE, FALSE), nrows)
isSpike(sce, "Spike") <- is.spike
sce
```

```{r, echo=FALSE, results='hide'}
# Save some memory.
rm(mito.data, endo.data, spike.data, new.counts)
gc()
```

# Quality control on the cells 

The original authors of the study have already removed low-quality cells prior to data publication.
Nonetheless, we compute some quality control metrics with `r Biocpkg("scater")` to check whether the remaining cells are satisfactory.

```{r}
library(scater)
sce <- calculateQCMetrics(sce, feature_controls=list(Spike=is.spike, Mt=is.mito)) 
```

We examine the distribution of library sizes and numbers of expressed genes across cells (Figure \@ref(fig:libplotbrain)).
In particular, the spike-in proportions here are more variable than in the HSC dataset.
This may reflect a greater variability in the total amount of endogenous RNA per cell when many cell types are present.

```{r libplotbrain, fig.wide=TRUE, fig.cap="Histograms of QC metrics including the library sizes, number of expressed genes and proportion of UMIs assigned to spike-in transcripts or mitochondrial genes for all cells in the brain dataset."}
par(mfrow=c(2,2), mar=c(5.1, 4.1, 0.1, 0.1))
hist(sce$total_counts/1e3, xlab="Library sizes (thousands)", main="", 
    breaks=20, col="grey80", ylab="Number of cells")
hist(sce$total_features, xlab="Number of expressed genes", main="", 
    breaks=20, col="grey80", ylab="Number of cells")
hist(sce$pct_counts_Spike, xlab="ERCC proportion (%)",
    ylab="Number of cells", breaks=20, main="", col="grey80")
hist(sce$pct_counts_Mt, xlab="Mitochondrial proportion (%)", 
    ylab="Number of cells", breaks=20, main="", col="grey80")
```

We remove small outliers for the library size and the number of expressed features, and large outliers for the spike-in proportions.
Again, the presence of spike-in transcripts means that we do not have to use the mitochondrial proportions.

```{r}
libsize.drop <- isOutlier(sce$total_counts, nmads=3, type="lower", log=TRUE)
feature.drop <- isOutlier(sce$total_features, nmads=3, type="lower", log=TRUE)
spike.drop <- isOutlier(sce$pct_counts_Spike, nmads=3, type="higher")
```

Removal of low-quality cells is then performed by combining the filters for all of the metrics.
The vast majority of cells are retained, which suggests that the original quality control procedures were generally adequate.

```{r}
sce <- sce[,!(libsize.drop | feature.drop | spike.drop)]
data.frame(ByLibSize=sum(libsize.drop), ByFeature=sum(feature.drop), 
    BySpike=sum(spike.drop), Remaining=ncol(sce))
```

```{r echo=FALSE, results='hide'}
gc()
```

# Cell cycle classification

Application of `cyclone` to the brain dataset suggests that most of the cells are in G1 phase (Figure \@ref(fig:phaseplotbrain)).
However, the intepretation of this result requires some caution due to the differences between the test and training datasets.
The classifier was trained on C1 SMARTer data [@scialdone2015computational] and accounts for the biases in that protocol. 
The brain dataset uses UMI counts, which has an entirely different set of biases, e.g., 3'-end coverage only, no length bias, no amplification noise.
These new biases (and the absence of expected biases) may interfere with accurate classification of some cells.

```{r phaseplotbrain, message=FALSE, fig.cap="Cell cycle phase scores from applying the pair-based classifier on the brain dataset, where each point represents a cell."}
library(org.Mm.eg.db)
ensembl <- mapIds(org.Mm.eg.db, keys=rownames(sce), keytype="SYMBOL", column="ENSEMBL")

library(scran)
mm.pairs <- readRDS(system.file("exdata", "mouse_cycle_markers.rds", package="scran"))
assignments <- cyclone(sce, mm.pairs, gene.names=ensembl)
plot(assignments$score$G1, assignments$score$G2M, xlab="G1 score", ylab="G2/M score", pch=16)
```

An additional complication is that many neuronal cell types are expected to lie in the G0 resting phase, which is distinct from the other phases of the cell cycle [@coller2006new].
Application of `cyclone` to these cells may be suboptimal if each cell must be assigned into one of the G1, S or G2/M phases.
To avoid problems from misclassification, we will not perform any processing of this dataset by cell cycle phase.
This is unlikely to be problematic here, as the cell cycle effect will be relatively subtle compared to the obvious differences between cell types in a diverse population.
Thus, the former is unlikely to distort the conclusions regarding the latter.

```{r echo=FALSE, results='hide'}
gc()
```

# Examining gene-level metrics

Figure \@ref(fig:topgenebrain) shows the most highly expressed genes across the cell population in the brain data set.
This is mostly occupied by spike-in transcripts, suggesting that too much spike-in RNA may be have been used.
There are also a number of constitutively expressed genes, as expected.

```{r topgenebrain, fig.asp=1.5, fig.wide=TRUE, fig.cap="Percentage of total counts assigned to the top 50 most highly-abundant features in the brain dataset. For each feature, each bar represents the percentage assigned to that feature for a single cell, while the circle represents the average across all cells. Bars are coloured by the total number of expressed features in each cell, while circles are coloured according to whether the feature is labelled as a control feature."}
fontsize <- theme(axis.text=element_text(size=12), axis.title=element_text(size=16))
plotQC(sce, type = "highest-expression", n=50) + fontsize
```

Gene abundance is quantified by computing the average count across all cells (Figure \@ref(fig:abhistbrain)).
As previously mentioned, the UMI count is generally lower than the read count.
This is because each transcript can only produce one UMI count but can yield many reads after fragmentation.
Some power will be lost due to the decrease in the size of the counts, but this is mitigated by a concomitant reduction in their variability.
Specifically, the use of UMIs eliminates technical noise due to amplification biases [@islam2014quantitative].

```{r abhistbrain, fig.cap="Histogram of log-average counts for all genes in the brain dataset. The filter threshold is represented by the blue line."}
ave.counts <- calcAverage(sce)
hist(log10(ave.counts), breaks=100, main="", col="grey",
    xlab=expression(Log[10]~"average count"))
abline(v=log10(0.1), col="blue", lwd=2, lty=2)
```

We save the average counts into the `SingleCellExperiment` object for later use.
We also remove genes that have average counts of zero, as this means that they are not expressed in any cell.

```{r}
rowData(sce)$ave.count <- ave.counts
to.keep <- ave.counts > 0
sce <- sce[to.keep,]
summary(to.keep)
```

```{r echo=FALSE, results='hide'}
gc()
```

# Normalization of cell-specific biases

For endogenous genes, normalization is performed using the deconvolution method in the `computeSumFactors` function.
Here, we cluster similar cells together and normalize the cells in each cluster using the deconvolution method.
This improves normalization accuracy by reducing the number of DE genes between cells in the same cluster.
Scaling is then performed to ensure that size factors of cells in different clusters are comparable.

```{r}
high.ave <- rowData(sce)$ave.count >= 0.1
clusters <- quickCluster(sce, subset.row=high.ave, method="igraph")
sce <- computeSumFactors(sce, cluster=clusters, 
    subset.row=high.ave, min.mean=NULL)
summary(sizeFactors(sce))
```

We set `subset.row` to use only the high-abundance genes for normalization (and clustering, for consistency).
We use a threshold of 0.1 to define `high.ave`, which is lower than the threshold in the HSC analysis to reflect the fact that UMI counts are generally smaller.
Setting `min.mean=NULL` simply avoids recomputing the average count within `computeSumFactors`, given that filtering has already been performed with `subset.row`.
(Here, `subset.row=high.ave` and `min.mean=0.1` are redundant as they do exactly the same thing.)

```{r echo=FALSE, results='hide'}
gc()
```

Compared to the HSC analysis, more scatter is observed around the trend between the total count and size factor for each cell (Figure \@ref(fig:normplotbrain)).
This is consistent with an increased amount of DE between cells of different types, which compromises the accuracy of library size normalization [@robinson2010scaling].
In contrast, the size factors are estimated based on median ratios and are more robust to the presence of DE between cells.

```{r normplotbrain, fig.cap="Size factors from deconvolution, plotted against library sizes for all cells in the brain dataset. Axes are shown on a log-scale."}
plot(sizeFactors(sce), sce$total_counts/1e3, log="xy",
    ylab="Library size (thousands)", xlab="Size factor")
```

We also compute size factors specific to the spike-in set, as previously described.

```{r}
sce <- computeSpikeFactors(sce, type="Spike", general.use=FALSE)
```

Finally, normalized log-expression values are computed for each endogenous gene or spike-in transcript using the appropriate size factors.

```{r}
sce <- normalize(sce)
```

```{r echo=FALSE, results='hide'}
gc()
```

__Comments from Aaron:__

- Only a rough clustering is required to avoid pooling together very different cell types in `computeSumFactors`.
The function is robust to a moderate level of differential expression between cells in the same cluster.
- For large data sets, using `method="igraph"` in `quickCluster` will speed up clustering. 
This uses a graph-based clustering algorithm - see `?buildSNNGraph` for more details.

# Checking for important technical factors

Larger experiments contain more technical factors that need to be investigated.
In this dataset, factors include the sex of the animal from which the cells were extracted, the age of the animal, the tissue of origin for each cell, and the total spike-in count in each cell.
Figure \@ref(fig:explvarplotbrain) shows that the tissue of origin explains a substantial proportion of the variance for a subset of genes.
This is probably because each tissue contains a different composition of cell types, leading to systematic differences in gene expression between tissues.
The other factors explain only a small proportion of the variance for most genes and do not need to be incorporated into our downstream analyses.

```{r, echo=FALSE, results='hide', message=FALSE}
fontsize <- theme(axis.text=element_text(size=12), axis.title=element_text(size=16))
```

```{r explvarplotbrain, fig.cap="Density plot of the percentage of variance explained by each factor across all genes in the brain dataset. For each gene, the percentage of the variance of the normalized log-expression values that is explained by the (log-transformed) total spike-in counts, the sex or age of the mouse, or the tissue of origin is calculated. Each curve corresponds to one factor and represents the distribution of percentages across all genes."}
plotExplanatoryVariables(sce, variables=c("log10_total_counts_Spike", 
    "log10_total_counts_Spike", "sex", "tissue", "age")) + fontsize
```

Nonetheless, we demonstrate how to account for uninteresting technical factors by using sex as an example.
We set up a design matrix with the sex of the animal as the explanatory factor for each cell.
This ensures that any sex-specific changes in expression will be modelled in our downstream analyses.
We do not block on the tissue of origin, despite the fact that it explains more of the variance than sex in Figure \@ref(fig:explvarplotbrain).
This is because the tissue factor is likely to be associated with genuine differences between cell types, so including it in the model might regress out interesting biological effects.

```{r}
design <- model.matrix(~sce$sex)
```

Other relevant factors include the chip or plate on which the cells were processed and the batch in which the libraries were sequenced.
Blocking on these factors may be necessary to account for batch effects that are often observed in scRNA-seq data [@hicks2015widespread; @tung2016batch].

# Modelling and removing technical noise

We model the technical noise by fitting a mean-variance trend to the spike-in transcripts, as previously described.
To account for uninteresting factors, we supply `design` to `trendVar` to regress out any technical differences due to sex.

```{r}
var.fit <- trendVar(sce, parametric=TRUE, span=0.4, design=design)
var.out <- decomposeVar(sce, var.fit)
```

Figure \@ref(fig:hvgplotbrain) indicates that the trend is fitted accurately to the technical variances.
The technical and total variances are also much smaller than those in the HSC dataset.
This is due to the use of UMIs which reduces the noise caused by variable PCR amplification.
Furthermore, the spike-in trend is consistently lower than the variances of the endogenous genes.
This reflects the heterogeneity in gene expression across cells of different types.
It also provides an example where most genes _are_ highly variable, such that fitting a trend to their variances would not recover the technical component.

```{r hvgplotbrain, fig.cap="Variance of normalized log-expression values against the mean for each gene, calculated across all cells in the brain data set after blocking on the sex effect. The blue line represents the mean-dependent trend in the technical variance of the spike-in transcripts (also highlighted as red points)."}
plot(var.out$mean, var.out$total, pch=16, cex=0.6, xlab="Mean log-expression", 
    ylab="Variance of log-expression")
points(var.out$mean[isSpike(sce)], var.out$total[isSpike(sce)], col="red", pch=16)
curve(var.fit$trend(x), col="dodgerblue", add=TRUE, lwd=2)
```

We check the distribution of expression values for the genes with the largest biological components to ensure that they are not driven by outliers (Figure \@(fig:hvgvioplotbrain)).
Some tweaking of the `plotExpression` parameters is necessary to visualize a large number of cells.

```{r hvgvioplotbrain, fig.cap="Violin plots of normalized log-expression values for the top 10 HVGs in the brain dataset. For each gene, each point represents the log-expression value for an individual cell."}
chosen.genes <- order(var.out$bio, decreasing=TRUE)[1:10]
plotExpression(sce, rownames(var.out)[chosen.genes], 
    alpha=0.05, jitter="jitter") + fontsize
```

Finally, we denoise the expression values using our PCA-based approach.
We supply `design` to regress out uninteresting factors, and we use only the genes with positive biological components, 
This yields a set of coordinates for each cell where the technical noise has been removed.

```{r}
sce <- denoisePCA(sce, technical=var.fit$trend, design=design, approximate=TRUE)
ncol(reducedDim(sce, "PCA"))
```

```{r, echo=FALSE, results='hide', message=FALSE}
gc()
```

__Comments from Aaron:__

- For data sets containing multiple batches, an alternative strategy is to perform trend fitting and variance decomposition separately for each batch.
This accommodates differences in the mean-variance trends between batches, especially if a different amount of spike-in RNA was added to the cells in each batch.
We demonstrate the second approach below by treating each sex as a different "batch".
Statistics are combined across multiple batches using the `combineVar` function.

```{r}
collected <- list()
for (block in levels(sce$sex)) {
    cur.sce <- sce[,sce$sex==block]
    cur.sce <- normalize(cur.sce) 
    var.fit <- trendVar(cur.sce, parametric=TRUE, span=0.4)
    collected[[block]] <- decomposeVar(cur.sce, var.fit)
}
var.out <- do.call(combineVar, collected)
```

```{r, echo=FALSE, results='hide', message=FALSE}
rm(cur.sce)
gc()
```

- Some downstream procedures must be performed on all batches at once, e.g., clustering or dimensionality reduction of all cells across multiple batches.
However, many of these procedures are not model-based and thus do not accept a design matrix to account for the batch effect.
To remove uninteresting factors of variation beforehand, we use the `removeBatchEffect` function from the `r Biocpkg("limma")` package [@ritchie2015limma].
This computes new expression values where the batch effect is regressed out, ensuring that it does not drive separation between clusters or in low-dimensional space.
This is demonstrated below for the sex effect in the brain data.
Note that this step is automatically performed inside `denoisePCA` when `design` is supplied, and does not need to be repeated.

```{r}
library(limma)
adj.exprs <- logcounts(sce)
adj.exprs <- removeBatchEffect(adj.exprs, batch=sce$sex)
norm_exprs(sce) <- adj.exprs 
```

```{r, echo=FALSE, results='hide', message=FALSE}
rm(adj.exprs)
gc()
```

- That being said, if an analysis method can accept a design matrix, blocking on nuisance factors in the design matrix is preferable to using `removeBatchEffect`.
This is because the latter does not account for the loss of residual degrees of freedom, nor the uncertainty of estimation of the blocking factor terms.
- Setting `approximate=TRUE` in `denoisePCA` will perform an approximate singular value decomposition, using methods from the `r CRANpkg("irlba")` package.
This is much faster than the exact algorithm on large data sets, without much loss of accuracy.

# Data exploration with dimensionality reduction

We perform dimensionality reduction on the denoised PCs to check if there is any substructure. 
Cells separate into clear clusters in the _t_-SNE plot (Figure \@ref(fig:tsneplotbrain)), corresponding to distinct subpopulations.
This is consistent with the presence of multiple cell types in the diverse brain population.

```{r tsneplotbrain, fig.cap="_t_-SNE plots constructed from the denoised PCs of the brain dataset. Each point represents a cell and is coloured according to its expression of _Neurod6_ (left) or _Mog_ (right).", fig.width=12, fig.asp=0.5}
sce <- runTSNE(sce, use_dimred="PCA", perplexity=10, rand_seed=100)
tsne1 <- plotTSNE(sce, colour_by="Neurod6") + fontsize
tsne2 <- plotTSNE(sce, colour_by="Mog") + fontsize
multiplot(tsne1, tsne2, cols=2)
```

The PCA plot is less effective at separating cells into many different clusters (Figure \@ref(fig:pcaplotbrain)).
This is because the first two PCs are driven by strong differences between specific subpopulations, which reduces the resolution of more subtle differences between some of the other subpopulations.
Nonetheless, some substructure is still visible.

```{r pcaplotbrain, fig.cap="PCA plots constructed from the denoised PCs of the brain dataset. Each point represents a cell and is coloured according to its expression of the _Neurod6_ (left) or _Mog_ (right).", fig.width=12, fig.asp=0.5}
pca1 <- plotReducedDim(sce, use_dimred="PCA", colour_by="Neurod6") + fontsize
pca2 <- plotReducedDim(sce, use_dimred="PCA", colour_by="Mog") + fontsize
multiplot(pca1, pca2, cols=2)
```

For both methods, we colour each cell based on the expression of a particular gene.
This is a useful strategy for visualizing changes in expression across the lower-dimensional space.
It can also be used to characterise each cluster if the selected genes are known markers for particular cell types.
For example, _Mog_ can be used to identify clusters corresponding to oligodendrocytes.

```{r echo=FALSE, results='hide'}
rm(tsne1, tsne2, pca1, pca2)
gc()
```

## Clustering cells into putative subpopulations

The denoised log-expression values are used to cluster cells into putative subpopulations.
Specifically, we perform hierarchical clustering on the Euclidean distances between cells, using Ward's criterion to minimize the total variance within each cluster.
This yields a dendrogram that groups together cells with similar expression patterns across the chosen genes.
An alternative approach is to cluster on a matrix of distances derived from correlations (e.g., as in `quickCluster`).
This is more robust to noise and normalization errors, but is also less sensitive to subtle changes in the expression profiles. 

```{r}
pcs <- reducedDim(sce, "PCA")
my.dist <- dist(pcs)
my.tree <- hclust(my.dist, method="ward.D2")
```

Clusters are explicitly defined by applying a dynamic tree cut [@langfelder2008defining] to the dendrogram.
This exploits the shape of the branches in the dendrogram to refine the cluster definitions, and is more appropriate than `cutree` for complex dendrograms.
Greater control of the empirical clusters can be obtained by manually specifying `cutHeight` in `cutreeDynamic`.

```{r}
library(dynamicTreeCut)
my.clusters <- unname(cutreeDynamic(my.tree, distM=as.matrix(my.dist), verbose=0))
```

We visualize the cluster assignments for all cells on the _t_-SNE plot in Figure \@(fig:tsneclusterbrain).
Adjacent cells are generally assigned to the same cluster, indicating that the clustering procedure was applied correctly.

```{r tsneclusterbrain, message=FALSE, fig.cap="_t_-SNE plot of the denoised PCs of the brain data set. Each point represents a cell and is coloured according to the cluster identity to which it was assigned."}
sce$cluster <- factor(my.clusters)
plotTSNE(sce, use_dimred="PCA", colour_by="cluster",
    perplexity=10, rand_seed=100) + fontsize
```

We check the separatedness of the clusters using the silhouette width (Figure ((silhouettebrain))).
Cells with large positive silhouette widths are closer to other cells in the _same_ cluster than to cells in _different_ clusters.
Conversely, cells with negative widths are closer to other clusters than to other cells in the cluster to which it was assigned.
Each cluster would ideally contain many cells with large positive widths, indicating that it is well-separated from other clusters.
This can be used to gauge the optimal parameter values (e.g., cut height, number of clusters) that maximize the separation between clusters.
For example, we could vary the cut height in `cutreeDynamic` to maximize the average silhouette width across all cells.

```{r silhouettebrain, message=FALSE, fig.cap="Barplot of silhouette widths for cells in each cluster. Each cluster is assigned a colour and cells with positive widths are coloured according to the colour of its assigned cluster. Any cell with a negative width is coloured according to the colour of the cluster that it is closest to. The average width for all cells in each cluster is shown, along with the average width for all cells in the data set."}
library(cluster)
clust.col <- scater:::.get_palette("tableau10medium") # hidden scater colours
sil <- silhouette(my.clusters, dist = my.dist)
sil.cols <- clust.col[ifelse(sil[,3] > 0, sil[,1], sil[,2])]
sil.cols <- sil.cols[order(-sil[,1], sil[,3])]
plot(sil, main = paste(length(unique(my.clusters)), "clusters"), 
    border=sil.cols, col=sil.cols, do.col.sort=FALSE) 
```

```{r echo=FALSE, results='hide'}
gc()
```

__Comments from Aaron:__

- Very heterogeneous datasets may yield a few large clusters on the first round of clustering.
It can be useful to repeat the variance modelling, denoising and clustering using only the cells within each of the initial clusters.
This can be achieved by subsetting `sce` according to a particular level of `my.clusters`, and re-applying the relevant functions on the subset.
Doing so may focus on a different set of genes that define heterogeneity _within_ an initial cluster, as opposed to those that define differences _between_ the initial clusters.
This would allow fine-scale structure within each cluster to be explored at greater resolution. 
For simplicity, though, we will only use the broad clusters corresponding to clear subpopulations in this workflow.
- For larger data sets, consider using `buildSNNGraph` and methods from the `r CRANpkg("igraph")` package to perform clustering.
This builds a shared-nearest-neighbour graph [@xu2015identification] in which cells are the nodes and edges are formed between cells that share nearest neighbours.
Clusters are then defined as highly connected communities of cells within this graph.
This is more efficient than forming a pairwise distance matrix for large numbers of cells.
Clustering parameters can be optimized by maximizing the modularity score for the formed clusters.

```{r}
snn.gr <- buildSNNGraph(sce, use.dimred="PCA")
gr.clusters <- igraph::cluster_fast_greedy(snn.gr)
table(gr.clusters$membership)
```

```{r echo=FALSE, results='hide'}
rm(snn.gr, gr.clusters)
gc()
```

# Detecting marker genes between subpopulations

Once putative subpopulations are identified by clustering, we can identify marker genes for each cluster using the `findMarkers` function.
This fits a linear model to the log-expression values for each gene using `r Biocpkg("limma")` [@ritchie2015limma].
The aim is to test for DE in each cluster compared to the others while blocking on uninteresting factors in `design`.
The top DE genes are likely to be good candidate markers as they can effectively distinguish between cells in different clusters.

```{r}
markers <- findMarkers(sce, my.clusters, design=design)
```

For each cluster, the DE results of the relevant comparisons are consolidated into a single output table.
This allows a set of marker genes to be easily defined by taking the top DE genes from each pairwise comparison between clusters.
For example, to construct a marker set for cluster 1 from the top 10 genes of each comparison, one would filter `marker.set` to retain rows with `Top` less than or equal to 10.
Other statistics are also reported for each gene, including the adjusted p-values (see below) and the log-fold changes relative to every other cluster.

```{r, echo=FALSE, results="hide"}
old.digits <- options()$digits
options(digits=3)
```

```{r}
marker.set <- markers[["1"]]
head(marker.set, 10)
```

```{r, echo=FALSE, results="hide"}
options(digits=old.digits)
```

We save the list of candidate marker genes for further examination.
The `overlapExprs` function may also be useful here, to prioritize candidates where there is clear separation between the distributions of expression values of different clusters.

```{r}
write.table(marker.set, file="brain_marker_1.tsv", sep="\t", quote=FALSE, col.names=NA)
```

We visualize the expression profiles of the top candidates to verify that the DE signature is robust.
Figure \@ref(fig:heatmapmarkerbrain) indicates that most of the top markers have strong and consistent up- or downregulation in cells of cluster 1 compared to some or all of the other clusters.
Thus, cells from the subpopulation of interest can be identified as those that express the upregulated markers and do not express the downregulated markers.

```{r heatmapmarkerbrain, fig.cap="Heatmap of mean-centred normalized and corrected log-expression values for the top set of markers for cluster 1 in the brain dataset. Column colours represent the cluster to which each cell is assigned, as indicated by the legend."}
top.markers <- marker.set$Gene[marker.set$Top <= 10]
top.exprs <- norm_exprs(sce)[top.markers,,drop=FALSE]
heat.vals <- top.exprs - rowMeans(top.exprs)
pheatmap(heat.vals, cluster_cols=my.tree,
    annotation_col=data.frame(Cluster=factor(my.clusters), row.names=colnames(sce)),
    annotation_colors=list(Cluster=setNames(clust.col, seq_along(unique(my.clusters)))))
```

Many of the markers in Figure \@ref(fig:heatmapmarkerbrain) are not uniquely up- or downregulated in the chosen cluster.
Testing for unique DE tends to be too stringent as it overlooks important genes that are expressed in two or more clusters.
For example, in a mixed population of CD4^+^-only, CD8^+^-only, double-positive and double-negative T cells, neither _Cd4_ or _Cd8_ would be detected as subpopulation-specific markers because each gene is expressed in two subpopulations.
With our approach, both of these genes will be picked up as candidate markers as they will be DE between at least one pair of subpopulations.
A combination of markers can then be chosen to characterize a subpopulation, which is more flexible than trying to find uniquely DE genes.

__Comments from Aaron:__

- To avoid problems with discreteness when modelling the mean-variance relationship, `findMarkers` will automatically partition the data into low- and high-abundance genes.
Empirical Bayes shrinkage is performed in each partition separately, prior to calculation of _p_-values using the shrunk variance estimates.
This ensures that discreteness does not affect the inferences for high-abundance genes, without needing to entirely discard the low-abundance genes.
- `findMarkers` can also be directed to find genes that are DE between the chosen cluster and _all_ other clusters.
This should be done by setting `pval.type="all"`, which defines the p-value for each gene as the maximum value across all pairwise comparisons involving the chosen cluster.
Combined with `direction="up"`, this can be used to identify unique markers for each cluster.
However, this is sensitive to overclustering, as unique marker genes will no longer exist if a cluster is split into two smaller subclusters.
- It must be stressed that the (adjusted) _p_-values computed here cannot be properly interpreted as measures of significance.
This is because the clusters have been empirically identified from the data.
`r Biocpkg("limma")` does not account for the uncertainty of clustering, which means that the _p_-values are much lower than they should be. 
This is not a concern in other analyses where the groups are pre-defined.
- The `SingleCellExperiment` object can also be easily transformed for use in other DE analysis methods.
For example, the `convertTo` function can be used to construct a `DGEList` for input into the `r Biocpkg("edgeR")` pipeline [@robinson2010edgeR].
This allows users to construct their own marker detection pipeline, though we find that direct use of `findMarkers` is usually sufficient.

```{r}
library(edgeR)
y <- convertTo(sce, type="edgeR")
```

# Additional comments

Having completed the basic analysis, we save the `SingleCellExperiment` object with its associated data to file.
This is especially important here as the brain dataset is quite large.
If further analyses are to be performed, it would be inconvenient to have to repeat all of the pre-processing steps described above.

```{r}
saveRDS(file="brain_data.rds", sce)
```

```{r, echo=FALSE, results='hide'}
gc()
```

# Software availability

All software packages used in this workflow are publicly available from the Comprehensive R Archive Network (https://cran.r-project.org) or the Bioconductor project (http://bioconductor.org).
The specific version numbers of the packages used are shown below, along with the version of the R installation.
Version numbers of all Bioconductor packages correspond to release version 3.5 of the Bioconductor project.
```{r, echo=FALSE, results='asis'}
if (!on.bioc) { 
    cat("Users can install all required packages and execute the workflow by following the instructions at https://www.bioconductor.org/help/workflows/simpleSingleCell.\n")
}
cat("The workflow takes less than an hour to run on a desktop computer with 8 GB of memory.\n")
```

```{r}
sessionInfo()
```

```{r, eval=on.bioc, echo=FALSE, results='hide'}
unlink(all.basenames)
```

# Author contributions

A.T.L.L. developed and tested the workflow on all datasets.
A.T.L.L. and D.J.M. implemented improvements to the software packages required by the workflow.
J.C.M. provided direction to the software and workflow development.
All authors wrote and approved the final manuscript.

# Competing interests

No competing interests were disclosed.

# Grant information

A.T.L.L. and J.C.M. were supported by core funding from Cancer Research UK (award no. A17197).
D.J.M. was supported by a CJ Martin Fellowship from the National Health and Medical Research Council of Australia.
D.J.M and J.C.M. were also supported by core funding from EMBL.

# Acknowledgements

We would like to thank Antonio Scialdone for helpful discussions, as well as Michael Epstein, James R. Smith and John Wilson-Kanamori for testing the workflow on other datasets.

# References

